---
title: 'Homework 3: Multinomial Logit Models'
author: "Peng Peng"
date: "3/6/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())

pkgs = c("dplyr", "purrr", "tidyr", "bayesm", "ggplot2", "data.table", "dummies")
lapply(pkgs, library, character.only = T)

data("margarine")

data = left_join(margarine$choicePrice, margarine$demos, by = "hhid")

```

# Exercise 1: Data Description
```{r}
# Average and dispersion of the product characteristics 
#--------------------------------------------------------------------

# A list of functions to apply across all columns 
x = data[, 3:12]
f = function(x){
  list(mean(x), sd(x))
}

# Transpose the result matrix and add product characteristics 
pr_char = data.frame(t(sapply(x, f))) %>%
    rename(mean = X1, sd = X2) %>% 
    cbind(names(data[, 3:12])) %>%
    print()

# Market share of each product-----------------------------------------------------

# Match names to choices 
product = names(data[, 3:12])

setnames(data, old = product, new = as.character(c(1:10)))

# Calculate market share of brands for Stk
share_stk = data %>% 
  select(2:8) %>%
  filter(choice < 7) %>% 
  group_by(choice) %>%
  summarise (n = n()) %>%
  mutate(share = n / sum(n)) %>% 
  mutate(product = product[3:8]) %>% 
  select(product, share) %>%
  print()

# Calculate market share of brands for Tub 
share_tub = data %>% 
  select(2, 9:12) %>%
  filter(choice > 6) %>% 
  group_by(choice) %>%
  summarise (n = n()) %>%
  mutate(share = n / sum(n)) %>% 
  mutate(product = c("PSS_Tub", "PPk_Tub", "PFl_Tub", "PHse_Tub")) %>%
  select(product, share) %>%
  print()

# Map observed attributes to choice----------------------------------------------

t1 = table(data$choice, data$Income) 
t1 = cbind(as.matrix(product), t1)

t2 = table(data$choice, data$Fam_Size)
t2 = cbind(as.matrix(product), t2)

t3 = table(data$choice, data$Fs3_4)
t3 = cbind(as.matrix(product), t3)

t4 = table(data$choice, data$Fs5.)
t4 = cbind(as.matrix(product), t4)

t5 = table(data$choice, data$college)
t5 = cbind(as.matrix(product), t5)

t6 = table(data$choice, data$whtcollar)
t6 = cbind(as.matrix(product), t6)
```


# Exercise 2: First Model
A conditional logit model could be used to model the effect of price on demand. The probability of household $i$ choosing product $j$ can be written as:
$$Pr_{ij} = \frac{exp(X_{ij}\beta)}{\sum_{j=1}^{m}exp(X_{ij}\beta)}, $$ where $x_ij$ is the price of the product facing each household. 

For identification purpose, we normalize $\alpha_1$ to 0. We divide the probability by $\frac{exp(X_{i1}\beta + \alpha_1)}{exp(X_{i1}\beta + \alpha_1)}$
we get:
$$Pr_{ij} = \frac{exp((X_{ij} - X_{i1})\beta + (\alpha_j - \alpha_1))}{\sum_{1}^{m}exp((X_{ik} - X_{i1})\beta + (\alpha_k - \alpha_1))}.$$

$$L = \prod_{i = 1}^{n} \prod_{j = 1}^{m}Pr_{ij}\mathbb{1}[j = j*]$$ where $j*$ is the actual choice made by $i$.
 Log-likelihood is thus: 
 $$\prod_{i = 1}^{n} \prod_{j = 1}^{m}log(Pr_{ij}\mathbb{1}[j = j*]) \\
 = \prod_{i = 1}^{n} \prod_{j = 1}^{m}log(\frac{exp((X_{ij} - X_{i1})\beta + (\alpha_j - \alpha_1))}{\sum_{1}^{m}exp((X_{ik} - X_{i1})\beta + (\alpha_k - \alpha_1))}$$

```{r}

#Likelihood function---------------------------------------
data = data
cl_ll = function(beta){
  x = data[, 3:12] - data[, 3]# set price 1 as reference and substract price 1 from all prices
  b = beta[1] # constant beta 
  alpha = beta[2:11] # alternative-specific constant 
  alpha[1] = 0  # set alpha_1 to zero 
  x_beta = x * b 
  alpha_choice = matrix(nrow = nrow(data), ncol = 1)
  x_beta_j = matrix(nrow = nrow(data), ncol = 1)
  alpha_t = matrix(rep(t(alpha), times = nrow(data)), ncol = ncol(t(alpha)), byrow = T)
  
  for (i in 1: nrow(data)){
    jstar  = data[i, "choice"]
    alpha_j = alpha[jstar]
    alpha_choice[i] = alpha_j
    x_beta_j[i] = x_beta[i, jstar]
  }
  numerator = exp(x_beta_j + alpha_choice)
  xbetak =  exp(x_beta + alpha_t)
  denominator = rowSums(xbetak)
  pr = numerator / denominator
  ll = log(pr)
  cl_ll = -sum(ll)
}

# Maxmize log-likelihood function
fit_cl = nlm( f=cl_ll, p = c(rep(0, times = 11)))
par_cl = fit_cl$estimate
print(par_cl)

# Interpret the coefficient on price-------------------------------
# For one unit increase in price, we would expect the household's utility to decrease. Compare to the reference product, product 2, 4, 6 give households lower utility and other products give households higher utility.
```

# Multinomial Logit Model 
Household $i$'s probability of choosing product $j$ is modelled as:
$$Pr_{ij} = \frac{X_{i}\beta_j}{\sum_{l=1}^{m} exp(X_i\beta_l)},$$ where $X_i$ is a batter of household characteristics including income, family size, college, white collar, retired. 

The likelihood function is $$L = \prod_{i = 1}^{n} \prod_{j = 1}^{m}Pr_{ij}\mathbb{1}[j = j*]$$ where $j*$ is the actual choice made by $i$.
 Log-likelihood is thus: 
 $$\prod_{i = 1}^{n} \prod_{j = 1}^{m}log(Pr_{ij}\mathbb{1}[j = j*]) \\
 = \prod_{i = 1}^{n} \prod_{l = 1}^{m}log(\frac{X_{i}\beta_j}{\sum_{l=1}^{m} exp(X_i\beta_l)})$$
```{r}

#Likelihood function ------------------------------

library(dummies)
indicator = dummy("choice", data = data)

x_i = as.matrix(data[, 13:19])

ll_mn = function(beta){
  beta = matrix(beta, nrow = 7, byrow = T)
  beta[, 1] = 0
  x_i_beta_j = x_i %*% beta
  ex = exp(x_i_beta_j)
  pr = t((apply(ex, 1, function(x) x / sum(x))))
  pr_choice = pr * indicator 
  pr_choice = rowSums(pr_choice)
  ll_mn =  -sum(log(pr_choice))
  return(ll_mn)
}

fit_mn = nlm(f = ll_mn, p = c(rep(0, times = 70)))
par_mn = data.frame(matrix(fit_mn$estimate, nrow = 7, byrow = T)) 
names(par_mn) = names(data[, 3:12])
rownames(par_mn) = names(data[, 13:19])
print(par_mn)

# Check with package 
library(nnet)

function_mn = multinom(choice ~ Income+ Fs3_4+Fs5. + Fam_Size + college+whtcollar + retired, data = data )
summary(function_mn)

# Interpret coefficien on income -------------------------------------
# Compared to reference product, higher income yields households product 2 yields households lower utility. Similarly, product 3 yields households higher utility, compared to reference product 1. 


```

# Marginal Effects
```{r}
# Conditional logit------------------

# Calculate probability for each i and j
x = data[, 3:12] - data[, 3]
b = par_cl[1]
alpha = par_cl[2:11] 
x_beta = x * b 
alpha_choice = matrix(nrow = nrow(data), ncol = 1)
x_beta_j = matrix(nrow = nrow(data), ncol = 1)
alpha_t = matrix(rep(t(alpha), times = nrow(data)), ncol = ncol(t(alpha)), byrow = T)
xbetak =  exp(x_beta + alpha_t)
denominator = rowSums(xbetak)
pr_ij = as.matrix(xbetak/denominator) 
pr = t(pr_ij) %*% pr_ij * (-b)
a = matrix(rep(colSums(pr_ij) * b,10), ncol=10 )
a = a * diag(10)

me_cl = data.frame((pr + a)/nrow(data))
print(me_cl)

# Multinomial logit-------------------------------------------
x_i = as.matrix(data[, c("Income", "Fs3_4", "Fs5.",  "college", "whtcollar", "retired")])
x_i = as.matrix(cbind(x_i, rep(1, times = nrow(data))))
beta = matrix(fit_mn$estimate, nrow = 7, byrow = T)
x_i_beta_j = x_i %*% beta
ex = exp(x_i_beta_j)
pr = t((apply(ex, 1, function(x) x / sum(x))))
beta_income = matrix(beta[1, ])

beta_bar = pr %*% beta_income 
beta_bar_large = matrix(rep(beta_bar, 10), ncol = 10 )
beta_j = matrix(rep(t(beta_income)), nrow(data), byrow = T, ncol = 10)
me_mn = data.frame(colSums(pr * (beta_j - beta_bar_large))/nrow(data))  
names(me_mn) = "ME_Income"
row.names(me_mn) = names(data[, 3:12])
print(me_mn)


# Interpret marginal effects 
# For conditional logit, if the price of product 1 increases, we would expect households to be less likely to purchase product 1, more likely to purchase other products. 
# For multinomial logit, if the income of household increases, we would expect households to be less likely to purchase product 1, 4, 5, 7, 10, and more likely to purchase the other products. 
```


# Mixed Logit and IIA 
```{r}
# Mixed logit model-------------------------------------------------------
x_ij = data[, 3:12]
w_i = as.matrix(data[, c("Income", "Fs3_4", "Fs5.",  "college", "whtcollar", "retired")])
w_i = as.matrix(cbind(rep(1, times = nrow(data)), w_i))
indicator = indicator 


# Likelihood function
ml_ll = function(b){
  beta = b[1]
  gamma = matrix(b[2:71], nrow = 7, ncol =10,  byrow = T)
  gamma[, 1] = 0
  x_i_j_beta = x_ij * beta
  w_i_r_j = w_i %*% gamma
  num = rowSums(exp((x_i_j_beta + w_i_r_j) * indicator))
  denom = rowSums(exp((x_i_j_beta + w_i_r_j)))
  pr = num / denom 
  ml_ll = -sum(log(pr))
  return(ml_ll)
}

# Maximize likelihood 
fit_ml = nlm(f = ml_ll, p = rep(0, times = 71))
par_ml = fit_ml$estimate
par_ml = data.frame(matrix(fit_ml$estimate[2:71], nrow = 7, byrow = T)) 
par_ml = rbind(par_ml, rep(fit_ml$estimate[1], times = 10, byrow = T))
names(par_ml) = names(data[, 3:12])
rownames(par_ml) = c(names(data[, 13:19]), "price")


#Test IIA------------------------------------------------

# Restrict data to choices not equal to 1 
d_res = data %>%
        filter(choice!=1)
x_ij_res = d_res[, 4:12]
w_i_res = as.matrix(d_res[, c("Income", "Fs3_4", "Fs5.",  "college", "whtcollar", "retired")])
w_i_res = as.matrix(cbind(w_i_res, rep(1, times = nrow(d_res))))
indicator = dummy("choice", data = d_res)

# LikelihooD function for the restricted model
ml_ll_res = function(b){
  beta = b[1]
  gamma = matrix(b[2:64], nrow = 7, ncol = 9,  byrow = T)
  gamma[, 1] = 0
  x_i_j_beta = x_ij_res * beta
  w_i_r_j = w_i_res %*% gamma
  num = rowSums(exp(x_i_j_beta + w_i_r_j) * indicator)
  denom = rowSums(exp((x_i_j_beta + w_i_r_j)))
  pr = num / denom 
  ml_ll_res = -sum(log(pr))
  return(ml_ll_res)
}


fit_ml_res = nlm(f = ml_ll_res, p = rep(0, times = 64))
par_ml_res = fit_ml_res$estimate
par_ml_res = data.frame(matrix(fit_ml_res$estimate[2:71], nrow = 7, byrow = T)) 
par_ml_res = rbind(par_ml_res, rep(fit_ml_res$estimate[1], times = 10, byrow = T))
names(par_ml_res) = names(data[, 4:12])
rownames(par_ml_res) = c(names(d_res[, 13:19]), "price")

# Compute test statistics 
beta_f = fit_ml$estimate
beta_r = fit_ml_res$estimate
L_f = ml_ll(beta_f)
L_r = ml_ll_res(beta_r)
MTT = -2 * (L_f - L_r) 
library(chi)
pchi(MTT, 57, lower.tail = F)

# Therefore, IIA property is violated. 
```

